---
title: "Introduction to phenoBayes"
author: "Your Name"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to phenoBayes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Welcome to phenoBayes

The `phenoBayes` package provides a robust, uncertainty-aware framework for extracting phenological metrics from species distribution models (specifically, `mgcv` GAM/BAM objects). By utilizing Bayesian posterior simulation (`mgcv::rmvn`), the package fully propagates the uncertainty of spatiotemporal splines into easily interpretable ecological metrics.

## Example: *Nebria brevicollis*
This classic carabid beetle is known for its summer dormancy (aestivation). Traditional phenological metrics (like First Appearance Date or percentiles) struggle with bimodal distributions. Let's see how `phenoBayes` handles it.

```{r eval=FALSE}
library(phenoBayes)
library(mgcv)

# 1. Load your fitted spatiotemporal BAM model
mod <- readRDS("Nebria_brevicollis_Final_GAMM.rds")

# 2. Define a prediction grid (e.g., the Warm South scenario)
newdata_warm <- data.frame(
  Yday = 1:365, 
  LogEffort = 0, 
  Latitude = 51.0, 
  TMAM = 11.5,
  fYear = factor("2008"), 
  SubSite = factor("Alice Holt")
)

# 3. Extract Metrics!
estimate_centroid(mod, newdata_warm, n_sims = 1000)
estimate_modality(mod, newdata_warm, n_sims = 1000)
estimate_duration(mod, newdata_warm, n_sims = 1000)
```

## Continuous Reaction Norms
Instead of plotting individual scenarios, you can use `plot_metric_gradient()` to visualize how a metric shifts across an entire environmental continuum!

```{r eval=FALSE}
# Plot Centroid Shift across the Latitudinal Gradient
plot_metric_gradient(mod, dataset, metric = "centroid", x_axis = "latitude")
```
